Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-08 VegOzone.tex
Comments by Ane

\section{Data}
\label{sec:data}

@@

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in an aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik, where as the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{svanhovd_researchstation}\\
  %\includegraphics[width=8.3cm]{IMG_7739.JPG}
  \includegraphics[width=8.3cm]{fig01}
  \caption{NIBIO Environment Centre Svanhovd close by the settlement of Svanvik, Norway. (a) Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk; (b) Clover in the Svanhovd ozone garden. The plants had to be secured against herbivores with a wire-mesh fence. The plants shown are approximately $6-12\,\unit{cm}$ in height.}
  \label{fig:svanhovd_research_station}
\end{figure}

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species of which one cultivar is more sensitive to ozone than the other. Cultivated species were, e.g. clover, tobacco, and beans. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

For the purpose of relating observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables (temperature, pressure, precipitation). All variables have been obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Ozone monitoring data has also been added to EBAS database operated by \citet{NILU_EBAS}.

Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) which occurred from July 12--29 and destroyed about $18000\,\unit{hectare}$ of forest \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

Hourly averaged $2\,\unit{m}$ temperatures below $-20\,\unit{^\circ C}$ were observed in winter (January/February), while high temperatures (above $20\,\unit{^\circ C}$) occurred in summer (July), though more regularly in 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April. 2019 saw a very cold November with temperatures regularly below $-10\unit{^\circ C}$.

More rain events with accumulated daily precipitation $\sum_d \mathrm{Precip}$ above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c). $\sum_d \mathrm{Precip} \ge 20\,\unit{mm}$ was only breached twice in 2018 and once in 2019.

In 2018, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) displays more sequentially high values in spring (May) and summer (July) compared to 2019, while June 2019 saw more consecutive high irradiance than 2018. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached only in June.

\begin{figure*}[t]
  %\includegraphics[width=12cm]{data_svanvik_2018_2019}
  \includegraphics[width=12cm]{fig02}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate missing ozone monitoring data. (a) Hourly averaged ozone concentration; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity it varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

@@

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}.}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- present\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^\dagger$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^\dagger$ Exclusive ozone monitoring growing seasons 2018/19 operated by NILU.} % Table Footnotes
\end{table}

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control \citep{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA) \citep{ACP:Inness2013, ACP:Inness2019}. The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $29.3\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution ground-level ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control (note: since May 31st 2020 through the CAMS Atmosphere Data Store (ADS)). The regional air quality model reanalysis ensemble (CAMS regional Ensemble) is based on nine European state-of-the-art numerical air quality models \citep{Copernicus_Regio_AIRQ}. The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millennium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near real time, but only former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite derived tropospheric column ozone (TC), while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMS Regional Ensemble. The quality with respect to observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

>>>>

%%
\section{Data}
\label{sec:data}

@@

\subsection{NIBIO Environment Centre Svanhovd and atmospheric monitoring}
\label{subsec:atmo_svanvik}
The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The center focuses on research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. The locations of the atmospheric monitoring site and the ozone garden are marked in an aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). Svanhovd is located south of the settlement of Svanvik and the border to Russia, the Pasvik river, lies only a few hundred meters to the east.\\

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{svanhovd_researchstation}\\
  %\includegraphics[width=8.3cm]{IMG_7739.JPG}
  \includegraphics[width=8.3cm]{fig01}
  \caption{NIBIO Environment Centre Svanhovd close by the settlement of Svanvik, Norway. (a) Atmospheric monitoring site and ozone garden have been marked. Aerial photography \copyright Norges Kartverk; (b) Clover in the Svanhovd ozone garden. The plants had to be secured against herbivores with a wire-mesh fence. The plants shown are approximately $6-12\,\unit{cm}$ high.}
  \label{fig:svanhovd_research_station}
\end{figure}

An ozone garden was installed in the course of our project at the NIBIO Environment Centre Svanhovd and was likely the northern-most ozone garden to date. An ozone garden consists of selected plant species which are more sensitive to ozone and display visible injuris. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

For the purpose of relating observed damage on ozone sensitive plants to local ozone concentrations, an ozone monitor had been installed during the growing seasons 2018/19 in addition to existing monitors of atmospheric key variables. All variables were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ}. Our ozone monitoring data has also been added to EBAS database operated by \citet{NILU_EBAS} and is thus openly accessable.
Long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.
All relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data was recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

 As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition started later than in 2018. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak ozone concentration. In contrast, ozone concentration only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.
Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in spring (May) and summer (July) 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the results of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June as naturally expected.

\begin{figure*}[t]
  %\includegraphics[width=12cm]{data_svanvik_2018_2019}
  \includegraphics[width=12cm]{fig02}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate periods without ozone monitoring data. (a) Hourly averaged ozone concentration; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we derive a ground-level ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated continously by NILU at Svanhovd, but no ozone measurements have been conducted until the start of our project. Due to probable changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors which fell under air quality regulations in Europe, we will cross-calibrate our data with long-term ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted ozone monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data are described in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which are not available at all times and spatially sparse, we also evaluate various ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

@@

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}.}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- present\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^\dagger$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^\dagger$ Exclusive ozone monitoring in growing seasons 2018/19 for the present study operated by NILU.} % Table Footnotes
\end{table}

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control \citep{NILU2003}. {\bf TODO: Redo plot with data from 2019 for Esrange and Pallas?}}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

Details of reanalysis data sets referred to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA) \citep{ACP:Inness2013, ACP:Inness2019}. The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $29.3\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution ground-level ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control (note: since May 31st 2020 through the CAMS Atmosphere Data Store (ADS)). The regional air quality model reanalysis ensemble (CAMS regional Ensemble) is based on nine European state-of-the-art numerical air quality models \citep{Copernicus_Regio_AIRQ}. The ensemble mean is at fairly high spatial and temporal resolution compared to the global reanalysis from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $3.9\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millennium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near real time, but only the former covers a period of sufficient length for climate analysis (2003 -- present). All reanalyses use different but, at that time, latest versions of the operational weather forecast system (OpenIFS) from ECMWF. They differ substantially in the assimilated observational ozone data. The MACC reanalysis assimilates only satellite derived tropospheric column ozone (TC), while CAMSRA also includes ozone profiles from satellite retrievals. Most importantly for our study, in situ observations from ozone near-surface station networks are only assimilated in the CAMS Regional Ensemble. The quality with respect to observations in northern Fennoscandia will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.
